---
title: "PSYC259- Getting Started"
author: "Amanda Mae Woodward"
date: "10/6/2020"
output: html_document
---
# Day 1: R & RStudio 

## Goals for Today:
- Explain what R Studio and R are and why they are beneficial to use 
- Course Expectations
- Identify the Four Windows present in R Studio
- Describe how to use an R Mark Down File 

## Goal #1 Explain what R Studio and R are and why they are beneficial to use 

### Why R? 
- Free!
- Lots of Packages that make analyses easier 
- No point and Click 
- **Really** Good Graphics
- Easy to reuse code and also share

### Why RStudio? (opinions) 
- Easy to see code, outputs, and data visualizations
- More Intuitive
- Help files present and see how data are imported

## Goal #2: Course Expectations

### Ask Questions
- Coding *is* hard.
- A little bit of struggling is okay, but this course should not lead to feelings of existential dread.
- Ask your peers, ask your TA, ask me
- Stack Overflow, R Cookbook
- Remember, coding is cumulative, so it's best to address questions quickly

### Try, Try, Try
- The only way to get better at coding is to practice. 
- You'll get practice in many ways: class examples, assignments, and the final project

### Learning Environment
- We're all here to learn, and we're at different levels
- It's okay to make mistakes
- I don't expect anyone to know everything about R (it's okay if you know only a little)

### There is no one way to code!

## Goal 3: Identify the Four Windows present in R Studio

###Getting to know the space
You may notice that we’re using something called R Studio. I like it more than just using the R shell. It highlights the code so code blocks, text, and functions are all different colors. (more intuitive to me)

###R Mark Down (.Rmd File)
- stored file
- saves variables
- for class, make sure everything you want to submit is in the .Rmd.

###Console:
- can use for quick math
- if you want to view data, type it here
- installing packages
- accessing help files

###Environment:
- list of stored information in your R Studio space
- can call these items by typing the name in the .Rmd file and pressing run or by typing in console and hitting enter
- “Import Data set” - allows you to manually import your datasets (can do this with code too, but sometimes easier)
- Environment too crowded? - use the clear button to empty it (will need to re run code to get items back)

###Files/Plots/Packages/Help/Viewer
- gives everything else
- Files: allows you to search for a file in your directory
- Plots: will show the plots as we make them
- Packages: In addition to commands built into R, packages contain additional functions contributed by other programmers (you can do this to!). This section tells you what packages you have access to
- Help file: typing ? in front of a function in the Console will open a Help file in this window
- Viewer: allows you to see HTML files 


## Goal #4 Describe how to use an R Mark Down File 
- .Rmd files let you use text and code simultaneously 
- R will treat anything you write in this file as text
- If you want it to read code, you need to place it in a code chunk (version dependent)
- To make a code chunk write 3 accents (`) (on a QWERTY key board, it is the button above the left tab)
- Then write {r} 

```{r}

```
- This highlighted area is called a "code chunk." You need to close the chunk as I did above
- You can name your code chunk by writing it's name with no spaces after the r
```{r firstChunk}

```
- Names can help you search for code chunks in longer files

You can also write code in a code chunk, but need to use quotes or #
```{r writingText}
"you can write anything you'd like in quotes"

# hashtags/pound signs will let you turn a whole line into text.
```

###Other uses
- you can include other components to emphasize points in your .Rmd file
- some changes won't be seen until you knit the file 

###Knitting the file
You can choose to knit your file into another format to share your work. 













