---
title: "Graphing Part 2"
author: "Amanda Mae Woodward"
date: "11/23/2020"
output: html_document
---

#Learning Outcomes
By the end of this class, students should be able to: 
- Model interactions in ggplot and ggeffects
- Specify inputs and outputs to create a Shiny app
- Create a Shiny App to visualize data from a data set

##Before we get started
Shiny apps are advanced. Even if we're focusing on "basics" the coding behind it is more advanced than what we covered last week. It's **ok** if you walk away from the lesson not knowing what I'm talking about. 

The plan is to **show** you what you can build to as you program in R. There are tutorials that are ~18 hours to complete. We can't do that in one class. I've included links that I use when making shiny apps that you may want to try. 

###Warm-up
Both ggeffects and shiny will allow us to use the skills we talked about last week (re making graphs). Before we get started, let's make sure that we remember how to make graphs in ggplot:

####Practice
1. Open the mtcars dataset. Create a scatterplot for the mpg and disp variables. Include gear as a third variable 
2. Create a histogram of the cyl column. 

###Learning Outcome #1: Model interactions in ggplot and ggeffects
Now that we remember a bit about how to make graphs, we'll talk about how to model interactions

**Creating a regression model**
We'll start with a regression model that contains an interaction with a categorical predictor. 
We're going to use the mtcars dataset for both examples.

```{r}
class(mtcars$cyl)
mtcars$cyl<- as.factor(mtcars$cyl)
lm1<-lm(mpg~cyl*hp, data=mtcars)
summary(lm1)
```

We can graph this using the group argument in ggplot: 
```{r}
ggplot(mtcars,aes(hp, mpg), group=cyl)+geom_point()+geom_smooth(method="lm", aes(color=cyl))+ theme_classic()
```

**Creating a regression to plot**
```{r}
lm2<- lm(mpg~hp*disp,data=mtcars)
summary(lm2)
```

Looking at the model output, we can see that all of our predictors, and their interaction, are significant. The way we graphed the categorical interaction won't work in this case. 

We can use a new package called ggeffects. You can use it to create graphs for regressions or mixed models. These graphs will show your marginal effects, or the predicted value of y when holding other variables constant. 

In ggeffects, it will automatically split continuous IVs into three sensible intervals with which to plot the interaction.

```{r}
library(ggeffects)
library(dplyr)
```
The function ggpredict will use the model output from the model we just made, and requires us to specify the variables that you would like to plot

You'll also notice that we're going to use a pipe to feed data into the plot
```{r}
ggpredict(lm2, terms = c("hp","disp")) %>% plot()
```

ggeffects runs ggplot in the background, so you can use all sorts of ggplot features with it, like this version:
```{r}
ggpredict(lm2, terms = c("hp", "disp")) %>% plot() + labs(title = "Nonsense Plot",
  x = "horsepower",
  y = "miles per gallon"
)
```
You'll notice that I used a different way to label the x and y axis than last week, AND we added a title. If you'd prefer to use "xlab" and "ylab", that's okay. You can also add a title using ggtitle(). 

You can also use the apa theme, but we need to load the jtools library
```{r}
library(jtools)
```

```{r}
ggpredict(lm2, terms = c("hp", "disp")) %>% plot() + labs(title = "Nonsense Plot",
  x = "horsepower",
  y = "miles per gallon"
)+ jtools::theme_apa()
```

####Learning Outcome 1:Practice
1. Open the diamonds dataset. Create a plot for the following model: price ~ clarity*table  
2. Create an additional plot for the following model: price~ y*z

###Learning Outcome #2: Create basic Shiny apps to visualize data
shiny is a package in R that allows us to create interactive graphs. You can create dashboards and place interactive graphs online. 

Today should be treated as an introduction. You can learn more here: https://shiny.rstudio.com/tutorial/ 

People make shiny apps in different ways, I'm going to use naming conventions and descriptions used by R Studio developers because it makes it easier to follow along. 

But first, we need to load the shiny package. 
```{r}
library(shiny)
```

Before we start coding, we need to talk about the different aspects of a shiny app: 
**User Interface:** We're programming the part that someone would see in the plot
- If you place it on the internet, this would be the portion that others would see 
**Server:** This is the background code to create what the user sees. It's what the computer runs in the background. (ie R code)  

The **User Interface** creates the webpage, and the **server** tells R how to put the pieces together.

Then, **shinyApp** combines the information in the User interface and the Server. We'll start with an example below to get an idea of the structure. 

Just like we started with ggplot, we'll start with a shell. 
```{r}
ui <- fluidPage()
server <- function(input, output) {}
shinyApp(ui = ui, server = server)
```
I am using ui to stand for the user interface. This is going to be what someone would see. Notice that the server is a function that we are designing, and we're giving it inputs and outputs. In this case, it is an empty function. 

Whatever we put in the fluidpage() function above is going to be present in our shiny app. 
```{r}
ui <- fluidPage("Amanda is making a Shiny App")
server <- function(input, output) {}
shinyApp(ui = ui, server = server)
```
You can see that whatever I write in fluid page, shows up in the app. When we want to add more to what other people will see, we need to specify different inputs and outputs.   

###Learning Outcome 2: Specifying Inputs and Outputs**
In this section, we will talk about different inputs and outputs you can use. 

**Creating a scale**
Making graphs is complex. We're going to start with making a scale input because it helps us understand how the input functions work.  

All inputs follow similar structures, so you will be able to apply this general format to other options.

The function to make a scale is sliderInput() and takes the following inputs: 
sliderInput(inputId, label, min, max, value)

Just looking at the input code (not in the shiny app), it will look something like this: 
```{r}
sliderInput(inputId="number", label="Pick a number", value=13, min=1, max=100)
```

When we put it in the ui/server code we used to make the app: 
```{r}
ui <- fluidPage(
  sliderInput(inputId = "number", 
    label = "Pick a number", 
    value = 13, min = 1, max = 100),
  plotOutput("hist")
)

server <- function(input, output) {}

shinyApp(ui = ui, server = server)
```
This code lets a person using the Shiny app pick a number between 1 and 100. We can then use this to create an output that will change based on people's responses. 

**Creating a graph**
In this example, we'll create a histogram that plots a random normal sample of size n (whatever the person picks).

Before we look at the shiny app code, I need to walk you through the code we'll use as an output. 
**Rnorm**
Before we get started, I need to go through some code we'll use to simulate, or make up, some data. 

We are going to be using a function called rnorm() which gives us a set of random numbers from a normal distribution. The default is to draw from a normal distribution with mean= 0 and sd=1.
```{r}
rnorm(100)
```
We can change the mean and standard deviation associated with our random distribution by including 
the following arguments:
rnorm(# of draws, mean=, sd= )
```{r}
rnorm(100, 1, .4)
```
 
You may notice that your random numbers are different from my random numbers. To make sure we get the same numbers, we can use a function called set.seed() to make sure we are using the same "random."
```{r}
set.seed(13)
```
**Note:** We will have the same numbers, as long as you don't run commands multiple times. If you rerun code, it will give you a new set of random numbers

We'll put this into a plot (which will let me show you a way to graph vectors not in a dataframe)
```{r}
ggplot(, aes(rnorm(100)))+ geom_histogram(fill="darksalmon")+theme_classic()
```

```{r}
ui <- fluidPage(
  sliderInput(inputId = "number", 
    label = "Pick a number", 
    value = 13, min = 1, max = 100),
  plotOutput("hist")
)

server <- function(input, output) {
  output$hist <- renderPlot({
    ggplot(, aes(rnorm(input$number)))+ geom_histogram(fill="darksalmon")+theme_classic()
  })
}

shinyApp(ui = ui, server = server)
```
####Learning Outcome 2: Practice
1. Create a shiny app that allows people to select a number between 1 and 5000. 
2. Add an output that creates a graph based on these numbers.

**Challenge:** Create a graph based on two numbers chosen by the user

###Learning Outcome 3: Use data to create a shiny App. 

For this graph, we're going to use built in R data and allow users to choose what they'd like to graph. To do this, we'll use a dropdown list as the input. (for other inputs: https://shiny.rstudio.com/images/shiny-cheatsheet.pdf)

Just like we did before, I'll show you what code we're adding to the ui side before we add it. 

We're going to use the diamonds dataset for this practice
```{r}
data(diamonds)
```


**User Interface Set up**
The first part we'll set up is the ui. We'll be making a drop down list. 

To make a dropdown list, we'll use the selectInputs function. It takes the following arguments:
selectInput(inputId, label (what other people see), choices(that people can choose from))
```{r}

ui<- fluidPage(
  titlePanel("diamonds data"),
  sidebarLayout(
    sidebarPanel(
  selectInput(inputId= "colorChoice", label="Select Diamond Color", choices= c("D"="D", "E"="E", "F"= "F", "G"="G", "H"="H", "I"="I", "J"="J"))),
  mainPanel(
plotOutput("colorgraph")
)
)
)
server <- function(input, output) {
  output$colorgraph <- renderPlot({
    ggplot(diamonds[diamonds$color==input$colorChoice,], aes(carat))+ geom_histogram(fill= "lightseagreen",bins= 50)+theme_classic()
  })
}


shinyApp(ui = ui, server = server)
```




